Computer Assisted Verbal Autopsy: Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022

### 2025-06-24 manuscript/manuscript.tex
Rewrite paper with feedback from CGPT

\abstract{
\textbf{Background:} Verbal autopsies (VAs) collect information on deaths occurring outside traditional healthcare settings to estimate representative causes of death (CODs). Current computer models assign CODs at population-level accuracy comparable to physicians, but perform poorly at the individual level, largely due to reliance on structured questionnaire data and neglect of narrative free text. Recently, the large language model ChatGPT-4 demonstrated human-level performance on professional and academic benchmarks. While ChatGPT-4 shows promise in COD assignment, its application to VA narratives has not yet been evaluated.

\textbf{Methods:} We analyzed 6,939 VA records from Sierra Leone (2019–2022) to compare four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, against physician-assigned CODs at both population and individual levels. GPT models used narratives, whereas InterVA-5 and InSilicoVA relied on questionnaires. CODs were grouped into 19, 10, and 7 categories for adult, child, and neonatal deaths. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Chance Corrected Concordance (PCCC) were used to assess population and individual level agreement with physician coding respectively, stratified by age and COD.

\textbf{Results:} GPT-4 outperformed all models overall (PCCC=0.61), followed by GPT-3.5 (0.56) and InSilicoVA/InterVA-5 (0.44). GPT-4 achieved the highest PCCC for adult and neonatal deaths (0.64/0.58), with GPT-3.5 for child deaths (0.54). Across ages, model performance increased from 1 month to 14 years ($\sim$0.10–0.75 PCCC) and declined from 15 to 69 years ($\sim$0.70–0.35). GPT-4, GPT-3.5, and InSilicoVA achieved the highest PCCC in 17, 9, and 4 of the 30 CODs, respectively. At the population level, all models achieved comparable CSMF accuracies (0.74–0.79).

\textbf{Conclusion:} All models performed similarly at the population level, but GPT models and InSilicoVA showed greater accuracy for specific CODs at the individual level. GPT models demonstrated improvements over InterVA-5 and InSilicoVA models. This study provides foundational evidence for integrating large language models into computer assisted VA to support physicians, reducing ill-defined codes and improving agreement in COD assignment.
}

@@

In 2019, 41 million people died prematurely from noncommunicable diseases every year, accounting for 74\% of all deaths globally \citep{worldhealthorganizationNonCommunicableDiseases2019}. While most of these deaths are preventable, effective intervention requires evidence-based resource allocation targeting high-risk populations \citep{benzigerGlobalBurdenDisease2016}. Reliable mortality counts and accurate Cause of Death (COD) data are therefore essential for guiding public health policy and reducing premature mortality \citep{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, in many low-income countries, civil registration and vital statistics systems remain incomplete. Fewer than half of all deaths are registered, and among these, only 8\% have an assigned COD \citep{worldhealthorganizationSCOREHealthData2021}. To address this gap, Verbal Autopsy (VA) has been employed as a scalable method for collecting mortality data and assigning likely CODs, particularly for deaths that occur outside of healthcare facilities, which account for more than half of all deaths in these settings \citep{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014,adairWhoDiesWhere2021}.

@@

Initially, 11,920 records from the HEAL-SL study \citep{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022} were collected from dual-coded EVA, where each record was randomly coded by two different physicians that assigned CODs as International Classification of Diseases Revision 10 (ICD-10) codes \citep{worldhealthorganizationICD10InternationalStatistical2011}. For each record, two codes were assigned by two different randomly selected physicians, where codes were evaluated for agreement using Central Medical Evaluation Agreement 10 (CMEA-10) codes. CMEA-10 groups a range of similar ICD-10 codes together, where if they are in agreement if they are within the same group \citep{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 2). When codes were not in agreement, a record enters the reconciliation phase, where the two physicians were provided reasoning and initial codes from each other to: (1) keep their initial code (2) assign the other physician's code or (3) assign a new code. If codes were not in agreement after the reconciliation phase, a record enters the adjudication phase, where a third senior physician evaluates both physicians' reasoning and codes before and after reconciliation, and assigns a final code based on their evaluation.

Since computer models were compared to physicians in this study, there was more certainty that COD assignments agreed by both physicians were representative of physician assignment than when they disagreed \citep{barnettComparativeAccuracyDiagnosis2019,morrisFactorsAssociatedPhysician2010,hsiaoFactorsAssociatedPhysician2012}. Thus, 6942 physician agreed records of the 11,920 total records were used. For better comparison, all codes were standardized to CGHR-10 codes (see Additional File 1) that generalized ICD-10 codes into 19, 10, and 7 categories for the adult (12 to 69 years), child (28 days to 11 years), and neonatal (under 28 days) age groups. After conversion, a final total of 6939 physician agreed records (3826 adult, 2636 child, and 477 neonatal) were used for modelling and performance evaluation. See Appendix \ref{app-methods-data} for further details on data preprocessing and Tables \ref{tab-data-cod} and \ref{tab-data-age} for COD and age range distributions of the physician agreed records.

@@

Four computer models were used to assign COD for each of the 6939 physician agreed records: GPT-3.5, GPT-4, InterVA-5, and InSilicoVA. InterVA-5 and InSilicoVA are widely used and studied standard statistical models \citep{murrayUsingVerbalAutopsy2014,otiVerbalAutopsyInterpretation2010,benaraEvaluationMethodsAssigning2023,chandramohanEstimatingCausesDeath2021,tungaVerbalAutopsyModels2021,leitaoComparisonPhysiciancertifiedVerbal2014,jhaAutomatedPhysicianAssignment2019} for COD assignment in VAs under the openVA framework \citep{liOpenVAToolkitVerbal2023}. InterVA-5 applies Bayesian probabilistic modelling \citep{bayesEssaySolvingProblem1958} using a set of standardized symptoms from reports and related conditional probabilities from medical experts to assign CODs based on the highest probability \citep{byassIntegratedApproachProcessing2019,byassStrengtheningStandardisedInterpretation2012}. InSilicoVA improves upon InterVA (e.g. comparable probabilities across individuals, measures of uncertainty, and inclusion of additional data sources) with a hierarchical Bayesian framework and Markov Chain Monte Carlo (MCMC) simulations \citep{brooksMarkovChainMonte1998,chibMarkovChainMonte2001,hanMarkovChainMonte2001} to incorporate multiple sources of uncertainty for assigning CODs based on the highest probability \citep{mccormickProbabilisticCauseofDeathAssignment2016b}. GPT-3.5 \citep{brownLanguageModelsAre2020} and GPT-4 \citep{openaiGPT4TechnicalReport2023} are LLMs that utilize deep neural networks with transformer architectures \citep{vaswaniAttentionAllYou2017} and reinforcement learning from human feedback \citep{ouyangTrainingLanguageModels2022,christianoDeepReinforcementLearning2017,stiennonLearningSummarizeHuman2020,wirthSurveyPreferencebasedReinforcement2017} to follow instructions from prompts and provide human-level responses, with known differences in GPT-4 possessing multimodal capabilites (e.g. image/voice input/output), more recent training data, and improved responses compared to ChatGPT-3 \citep{wuBriefOverviewChatGPT2023}.

@@

\noindent For InterVA-5 and InSilicoVA, the standardized questionnaire data from EVA were converted into OpenVA format \citep{liOpenVAToolkitVerbal2023}, before being used as input for each model to produce COD assignments as WHO VA 2016 codes \citep{worldhealthorganizationVerbalAutopsyStandards2016}. All model outputs were converted to CGHR-10 codes to evaluate performances of models for COD assignment relative to physicians. See Appendix \ref{app-methods-model} for additional details regarding input parameters, output data, and code conversions for each model.

@@

The performance of the four models were evaluated with metrics at the population and individual level by comparing their CGHR-10 COD outputs for 6939 records. Cause Specific Mortality Fraction (CSMF) accuracy was used to evaluate models on the population level (see Appendix \ref{app-methods-csmfa}), while Partial Chance Corrected Concordance (PCCC) was used to evaluate models on the individual level (see Appendix \ref{app-methods-pccc}) \citep{murrayRobustMetricsAssessing2011}. Both CSMF accuracy and PCCC metrics are between 0 and 1 with 0 indicating low performance and 1 indicating perfect performance at the population and individual level respectively. As model performance can vary across ages and specific causes \citep{benaraEvaluationMethodsAssigning2023,murrayUsingVerbalAutopsy2014,setelValidityVerbalAutopsy2006}, the CSMF accuracy and PCCC metrics were compared for each model overall, by age group (adult, child, neonatal), by CGHR-10 COD codes, and across ages. For each of the adult and child age groups, metrics were calculated for five-year ages for records with ages at death of one-year or older and five-month ages for 28 days or older. For the neonatal age group, the ages of 0-6 days and 7-27 days were used. See Appendix \ref{app-methods-eval} for more details on performance metrics and evaluation strategy for comparing each model.

@@

This section details the performance results of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models for assigning CGHR-10 CODs after applying the methods in Section \ref{methods}. GPT-4 performed the best overall at 0.61 PCCC followed by GPT-3.5 at 0.56 PCCC. GPT-4 also had the highest PCCC for most ages and CODs across the adult (12 to 69 years), child (28 days to 11 years), and neonatal (under 28 days) age groups with GPT-3.5, InterVA-5, and InSilicoVA having higher PCCC values for a few ages and CODs. Overall performance results are seen in Section \ref{results-overall}, and performance by adult, child, and neonatal records are seen in Sections \ref{results-adult}, \ref{results-child}, and \ref{results-neo} respectively.

@@

Of all 6939 records, GPT-4 (0.61 PCCC) had the highest individual performance followed by GPT-3.5 (0.56 PCCC), InSilicoVA (0.44 PCCC), and InterVA-5 (0.44 PCCC) (Figure \ref{fig-pccc-csmf}). GPT-3.5 and GPT-4 had improvements ranging from 0.14-0.18 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.05 PCCC. Population level performances were similar for all models (0.74-0.79 CSMF). Figure \ref{fig-pccc} shows the PCCC performance across three age groups (adult, child, and neonate). GPT-4 had the best individual performance for adult and neonatal records (0.64 and 0.58 PCCC), while GPT-3.5 had the best performance for child records (0.54 PCCC) with GPT-4 performing slightly worse (0.51 PCCC). InSilicoVA and InterVA-5 performed the worse for adult and child records ($\leq$0.5 PCCC), while GPT-3.5 performed the worse for neonatal records (0.42 PCCC). Across ages, all models followed a similar pattern in individual performance (Figure \ref{fig-pccc-age}), where PCCC trended upwards for 1 month to 14 years ($\sim$0.1-0.75), and downwards for ages 15 to 69 years ($\sim$0.7-0.35). The highest and lowest performances were observed for ages 12-29 years ($\sim$0.4-0.7) and 1-11 months ($\sim$0.1-0.35) respectively.

@@

Figure \ref{fig-pccc-cod-adult} shows model performance by PCCC across 17 adult CODs excluding suicide due to low sample size (n=3, \textless 1\%). GPT-4 had the highest individual performance for 10 of 17 CODs (0.35 to 0.99 PCCC), GPT-3.5 for 5 CODs (0.43-0.94 PCCC), and InSilicoVA for 2 CODs (0.71 and 0.84 PCCC). InterVA-5 had the lowest performance for 8 of 17 CODs (0-0.79 PCCC), InSilicoVA for 6 CODs (0.01-0.41 PCCC), and GPT-3.5 for 2 CODs (0.38 and 0.53 PCCC). GPT-3.5/4 models improved over InSilicoVA/InterVA-5 the most for chronic respiratory diseases (0.74-0.94 PCCC difference), and the least for Malaria (0.09-0.17 PCCC difference). All models had \textgreater 0.7 PCCC for maternal conditions (0.79-0.99 PCCC), while \textless 0.5 PCCC for unspecified infections, malaria, and ill-defined CODs. GPT-4 had performance improvements \textgreater 0.2 PCCC compared to all other models for cancers (+0.25-0.36 PCCC), stroke (+0.27-0.45 PCCC), and diarrhoeal diseases (+0.37-0.51 PCCC), while GPT-3.5 had similar improvements for liver and alcohol related diseases (+0.27-0.52 PCCC).

@@

Figure \ref{fig-pccc-cod-child} presents individual performances for each of the models by 8 child CODs, excluding congenital anomalies due to low sample size (n=1, \textless 1\%). GPT-4 had the highest individual performance for 4 of 8 CODs (0.65-0.94 PCCC), GPT-3.5 for 3 CODs (0.44-0.88 PCCC), and InSilicoVA for 1 COD (0.78 PCCC). InterVA-5 had the lowest performance for 4 of 8 CODs (0.09-0.79 PCCC), InSilicoVA for 3 CODs (0-0.35 PCCC), and GPT-3.5 for 1 COD (0.58 PCCC). All models had \textgreater 0.7 PCCC for injuries (0.79-0.94 PCCC), and \textless 0.6 PCCC for malaria (0.35-0.54 PCCC) and other infections (0.29-0.44 PCCC). GPT-4 had improvements \textgreater 0.3 PCCC compared to other models for ill-defined CODs (+0.38-0.65 PCCC), and larger improvements over other models for injuries (+0.11-0.15 compared to +0.01-0.04 PCCC).

@@

Model performance across 5 neonatal CODs, excluding congenital anomalies (n=2, \textless 1\%) and other (n=5, 1\%) due to small sample sizes is shown in Figure \ref{fig-pccc-cod-neo}. GPT-4 had the highest individual performance for 3 of 5 CODs (0.39-0.71 PCCC), GPT-3.5 for 1 COD (0.57 PCCC), and InSilicoVA for 1 COD (0.86 PCCC). GPT-3.5 had the lowest performance for 3 of 5 CODs (0-0.13 PCCC) and InterVA-5 for 2 CODs (0.01 and 0.48 PCCC). All models had similar performance for stillbirth deaths (0.48-0.57 PCCC), while only GPT-4 had a PCCC \textgreater 0 PCCC. InSilicoVA had improvements over all other models for neonatal infection deaths (+0.18-0.73 PCCC).

@@

This section discusses and summarizes the results from Section \ref{results}. Advantages and disadvantages of using GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models for assigning CODs are discussed in Sections \ref{discuss-adv} and \ref{discuss-disadv}. Limitations of the study are mentioned in Section \ref{discuss-limits}, while opportunities and future work are detailed in Section \ref{discuss-opp}.

@@

This section identifies the advantages of models for assigning CODs. Section \ref{discuss-adv-cod} details the application of models for particular CODs and ages. Section \ref{discuss-adv-hscav} details the resource efficiency of computer models for assisting in physician COD assignment. Section \ref{discuss-adv-nl} notes the strength of using natural language text in GPT models compared to structured questionnaire data for physician COD assignment.

@@

At the population level, overall performances for all models were similar to physicians (0.74-0.79 CSMF), indicating potential for adequately estimating COD distributions for large populations. Although all models did not perform well for all records at the individual level (0.44-0.61 PCCC), several models performed well for certain CODs (0-0.99 PCCC). For most CODs, GPT-3.5/GPT-4 performed better than InSilicoVA/InterVA-5 (top PCCC for 15 of 17, 7 of 8, and 4 of 5 adult, child, and neonatal CODs respectively), while InSilicoVA performed better for particular CODs (road and transport injuries, tuberculosis, pneumonia, and neonatal infections with 0.84, 0.71, 0.78, and 0.86 PCCC respectively). For CODs with high performance (e.g. GPT-3.5/4 with 0.91-0.99 PCCC for maternal conditions, chronic respiratory disease, diabetes melitus, and cancers, InSilicoVA with 0.84 and 0.86 PCCC for road and transport injuries, and neonatal infections), the results suggest that GPT-3.5/4 and InSilicoVA may assign CODs that are very similar to physicians. Thus, it may be beneficial to evaluate performance at the COD level, and apply a combination of models that perform well in comparison to physicians for each COD. For example, different models perform well for various leading CODs as seen in Table \ref{tab-leadcod} \citep{carshon-marshChildMaternalAdult2022,ansumanaReportCausesDeath2023}.

@@

Across ages, all models followed a similar upward trend from 6 months to 14 years of age, and a downward trend from 15-69 years with GPT models having higher performance than InSilicoVA/InterVA-5 models, while more mixed trends were observed from 0 days to 5 months (recall Figure \ref{fig-pccc-age}). For adult ages, performance generally decreased as age increased, which suggested that models had difficult assigning CODs for older than younger adults with some improvements after the age of 59. For child and neonatal ages, the performance improved drastically as the age increased after 5 months, suggesting less difficulty in COD assignment when children and neonates are more developed. As the models did not perform particularly well ($\geq$ 0.8 PCCC) for any specific five-year age range, it is not recommended to apply specific models that target cases by age. However, the patterns of increases and decreases of performance in relation to age provide valuable insight for comparison to expected physician diagnosis patterns in well-studied medical literature and knowledge. For example, it may be expected that physicians are more uncertain in diagnosing diseases that are prevalent in neonatal patients \citep{rasmussenComplexityPhysiciansUnderstanding2019,faisonWhenUnknownUnknowable2023}, which are present in our findings from Figure \ref{fig-pccc-age}.

>>>>

\abstract{
\textbf{Background:} Verbal autopsies (VAs) collect information on deaths occurring outside traditional healthcare settings to estimate representative Causes of Death (CODs). Current computer models assign CODs at population-level accuracy comparable to physicians, but perform poorly at the individual level, largely due to reliance on structured questionnaire data and neglect of narrative free text. Recently, the large language model ChatGPT-4 demonstrated human-level performance on professional and academic benchmarks. While ChatGPT-4 shows promise in COD assignment, its application to VA narratives has not yet been evaluated.

\textbf{Methods:} We analyzed 6,939 VA records from Sierra Leone (2019–2022) to compare four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, against physician-assigned CODs at population and individual levels. GPT models used narratives, whereas InterVA-5 and InSilicoVA relied on questionnaires. CODs were grouped into 19, 10, and 7 categories for adult, child, and neonatal deaths. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Chance Corrected Concordance (PCCC) were used to assess population and individual level agreement with physician coding respectively, stratified by age and COD.

\textbf{Results:} GPT-4 outperformed all models overall (PCCC=0.61), followed by GPT-3.5 (0.56) and InSilicoVA/InterVA-5 (0.44). GPT-4 achieved the highest PCCC for adult and neonatal deaths (0.64 and 0.58), with GPT-3.5 for child deaths (0.54). Across ages, model performance increased from 1 month to 14 years ($\sim$0.10–0.75 PCCC) and declined from 15 to 69 years ($\sim$0.70–0.35). GPT-4, GPT-3.5, and InSilicoVA achieved the highest PCCC in 17, 9, and 4 of the 30 CODs, respectively. At the population level, all models achieved comparable CSMF accuracies (0.74–0.79).

\textbf{Conclusion:} All models performed similarly at the population level, but GPT models and InSilicoVA showed greater performance for specific CODs at the individual level. GPT models demonstrated improvements over InterVA-5 and InSilicoVA models. This study provides foundational evidence for integrating computer models to assist physicians with alternative diagnoses, helping reduce ill-defined codes and improve agreement in COD assignment.
}

@@

Every year, 41 million people died prematurely from noncommunicable diseases, accounting for 74\% of all deaths globally \citep{worldhealthorganizationNonCommunicableDiseases2019}. While most of these deaths are preventable, effective intervention requires evidence-based resource allocation that targets high-risk populations \citep{benzigerGlobalBurdenDisease2016}. Reliable mortality counts and accurate Cause of Death (COD) data are essential for guiding public health policy and reducing premature mortality \citep{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, in many low-income countries, civil registration and vital statistics systems remain incomplete. Fewer than half of all deaths are registered, and among these, only 8\% have an assigned COD \citep{worldhealthorganizationSCOREHealthData2021}. To address this gap, Verbal Autopsy (VA) has been deployed as a scalable method for collecting mortality data and assigning likely CODs, particularly for deaths that occur outside of healthcare facilities, which account for more than half of all deaths \citep{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014,adairWhoDiesWhere2021}.

@@

A total of 11,920 verbal autopsy (VA) records were obtained from the HEAL-SL study \citep{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022}, which employed dual-coded Electronic Verbal Autopsy (EVA). Each record was independently reviewed by two randomly selected physicians, who assigned COD codes based on the International Classification of Diseases, 10th Revision (ICD-10) \citep{worldhealthorganizationICD10InternationalStatistical2011}. Agreement between physician-assigned CODs was evaluated using Central Medical Evaluation Agreement 10 (CMEA-10) codes, which group related ICD-10 codes into broader, clinically similar categories \citep{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 2). If both codes fell within the same CMEA-10 group, the record was considered in agreement. Disagreements entered a reconciliation phase, where each physician was shown both the assigned codes and the reasoning from the other physician. Physicians could then (1) retain their original code, (2) adopt the other physician’s code, or (3) assign a new code. Records that remained unresolved proceeded to adjudication, where a senior physician reviewed all reasoning and assignments and issued a final COD.

To ensure comparability with physician coding, only records with physician agreement were used in this study, as such cases provide higher confidence in the COD assignment \citep{barnettComparativeAccuracyDiagnosis2019,morrisFactorsAssociatedPhysician2010,hsiaoFactorsAssociatedPhysician2012}. From the original dataset, 6,942 records met this criterion. All ICD-10 codes were then standardized to CGHR-10 categories (see Additional File 1), which group causes into 19, 10, and 7 categories for adults (12–69 years), children (28 days to 11 years), and neonates (under 28 days), respectively. After excluding three records without a valid CGHR-10 category, a total of 6,939 physician-agreed records (3,826 adult, 2,636 child, and 477 neonatal) were used for model comparison and performance evaluation. Further details on data preprocessing are provided in Appendix \ref{app-methods-data}, with COD and age group distributions summarized in Tables \ref{tab-data-cod} and \ref{tab-data-age}.

@@

Four computational models were used to assign causes of death (CODs) for each of the 6,939 physician-agreed verbal autopsy (VA) records: GPT-3.5, GPT-4, InterVA-5, and InSilicoVA. InterVA-5 and InSilicoVA are widely used statistical models within the OpenVA framework for COD assignment in VAs \citep{murrayUsingVerbalAutopsy2014, otiVerbalAutopsyInterpretation2010, benaraEvaluationMethodsAssigning2023, chandramohanEstimatingCausesDeath2021, tungaVerbalAutopsyModels2021, leitaoComparisonPhysiciancertifiedVerbal2014, jhaAutomatedPhysicianAssignment2019, liOpenVAToolkitVerbal2023}. InterVA-5 applies a Bayesian probabilistic approach, using a standardized set of symptoms and expert-derived conditional probabilities to assign the most likely COD based on maximum probability \citep{byassIntegratedApproachProcessing2019, byassStrengtheningStandardisedInterpretation2012, bayesEssaySolvingProblem1958}. InSilicoVA extends this approach by incorporating a hierarchical Bayesian framework and Markov Chain Monte Carlo (MCMC) methods \citep{brooksMarkovChainMonte1998, chibMarkovChainMonte2001, hanMarkovChainMonte2001}, allowing for quantification of uncertainty, individual-level probability estimates, and the integration of additional data sources \citep{mccormickProbabilisticCauseofDeathAssignment2016b}.GPT-3.5 \citep{brownLanguageModelsAre2020} and GPT-4 \citep{openaiGPT4TechnicalReport2023} are large language models (LLMs) based on transformer architectures \citep{vaswaniAttentionAllYou2017}. These models are trained using reinforcement learning from human feedback \citep{ouyangTrainingLanguageModels2022, christianoDeepReinforcementLearning2017, stiennonLearningSummarizeHuman2020, wirthSurveyPreferencebasedReinforcement2017}, enabling them to follow natural language instructions and generate human-level responses. GPT-4 introduces improvements over GPT-3.5, including more recent training data, enhanced reasoning capabilities, and multimodal input-output functionality (e.g. text, image, voice) \citep{wuBriefOverviewChatGPT2023}.

@@

\noindent InterVA-5 and InSilicoVA used structured questionnaire data, which were converted into OpenVA-compatible format \citep{liOpenVAToolkitVerbal2023}. Both models produced COD assignments coded using the WHO 2016 VA standard \citep{worldhealthorganizationVerbalAutopsyStandards2016}. To ensure comparability across models, all output CODs were mapped to the CGHR-10 classification system for evaluation relative to physician-assigned CODs. Further details on model input formats, output mappings, and code conversion procedures are provided in Appendix \ref{app-methods-model}.

@@

Model performance was assessed at both the population and individual levels by comparing each model’s CGHR-10 COD assignments to those of physicians for all 6,939 records. Cause-Specific Mortality Fraction (CSMF) accuracy was used to evaluate agreement at the population level (see Appendix \ref{app-methods-csmfa}), while Partial Chance-Corrected Concordance (PCCC) was used to assess individual-level agreement (see Appendix \ref{app-methods-pccc}) \citep{murrayRobustMetricsAssessing2011}. Both metrics range from 0 to 1, where higher values indicate stronger similarity with physician assignment.

Given that model performance can vary by age and different CODs \citep{benaraEvaluationMethodsAssigning2023, murrayUsingVerbalAutopsy2014, setelValidityVerbalAutopsy2006}, both CSMF accuracy and PCCC were calculated overall and stratified by age group (adult, child, neonatal), CGHR-10 COD, and age at death. For adult and child groups, metrics were computed in five-year age bands for records with age at death of one year or older, and five-month bands for records between 28 days and one year. For the neonatal group, evaluations were conducted separately for age intervals of 0–6 days and 7–27 days. Additional details on the evaluation strategy and metric calculations are provided in Appendix \ref{app-methods-eval}.

@@

This section presents the performance of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA in assigning CGHR-10 CODs, based on the methodology described in Section \ref{methods}. GPT-4 achieved the highest overall individual-level concordance, with a PCCC of 0.61, followed by GPT-3.5 (0.56). GPT-4 also demonstrated the highest PCCC across most age groups and CODs within the adult (12–69 years), child (28 days–11 years), and neonatal (under 28 days) categories. In contrast, GPT-3.5, InterVA-5, and InSilicoVA showed higher PCCC values for a limited subset of age groups and CODs. Summary results are presented in Section \ref{results-overall}, with stratified results by age group detailed in Sections \ref{results-adult}, \ref{results-child}, and \ref{results-neo}.

@@

Of all 6939 records, GPT-4 (0.61 PCCC) had the highest individual performance followed by GPT-3.5 (0.56 PCCC), InSilicoVA (0.44 PCCC), and InterVA-5 (0.44 PCCC) (Figure \ref{fig-pccc-csmf}). GPT-3.5 and GPT-4 had improvements ranging from 0.14-0.18 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.05 PCCC. Population level performances were similar for all models (0.74-0.79 CSMF). Figure \ref{fig-pccc} shows the PCCC performance across three age groups (adult, child, and neonate). GPT-4 had the best individual performance for adult and neonatal records (0.64 and 0.58 PCCC), while GPT-3.5 had the best performance for child records (0.54 PCCC) with GPT-4 performing slightly worse (0.51 PCCC). InSilicoVA and InterVA-5 performed the worse for adult and child records ($\leq$0.5 PCCC), while GPT-3.5 performed the worse for neonatal records (0.42 PCCC). Performance varied less for child deaths (0.13 range) than for adult and neonatal deaths (0.24 and 0.22 range). Across ages, all models followed a similar pattern in individual performance (Figure \ref{fig-pccc-age}), where PCCC trended upwards for 1 month to 14 years ($\sim$0.1-0.75), and downwards for ages 15 to 69 years ($\sim$0.7-0.35). The highest and lowest performances were observed for ages 12-29 years ($\sim$0.4-0.7) and 1-11 months ($\sim$0.1-0.35) respectively.

@@

Figure \ref{fig-pccc-cod-adult} presents model performance across 17 adult CODs, excluding suicide due to a low sample size (n=3, \textless 1\%). GPT-4 achieved the highest individual level performance for 10 of 17 CODs (0.35–0.99 PCCC), followed by GPT-3.5 for 5 CODs (0.43–0.94 PCCC), and InSilicoVA for 2 CODs (0.71 and 0.84 PCCC). InterVA-5 showed the lowest performance for 8 CODs (0–0.79 PCCC), InSilicoVA for 6 CODs (0.01–0.41 PCCC), and GPT-3.5 for 2 CODs (0.38 and 0.53 PCCC). The greatest improvements of GPT-3.5/4 over InSilicoVA and InterVA-5 were observed in chronic respiratory diseases (+0.74-0.94 PCCC), while the smallest improvements were for malaria (+0.09-0.17 PCCC). All models achieved PCCC values above 0.70 for maternal conditions (0.79–0.99), but remained below 0.50 for unspecified infections (0.35–0.49), malaria (0.26–0.43), and ill-defined CODs (0–0.35). GPT-4 showed performance improvements exceeding 0.20 PCCC over all other models for cancers (+0.25–0.36), stroke (+0.27–0.45), and diarrhoeal diseases (+0.37–0.51). GPT-3.5 demonstrated similar gains for liver and alcohol-related diseases (+0.27–0.52). Performance variability across models was most pronounced for chronic respiratory diseases (range: 0.94), while narrower differences were observed for maternal conditions (0.20), malaria (0.17), ischemic heart disease (0.15), and unspecified infections (0.14).

@@

Figure \ref{fig-pccc-cod-child} shows individual-level performance across 8 child CODs, excluding congenital anomalies due to a low sample size (n=1, \textless 1\%). GPT-4 achieved the highest PCCC for 4 of the 8 CODs (0.65–0.94), followed by GPT-3.5 for 3 CODs (0.44–0.88), and InSilicoVA for 1 COD (0.78). InterVA-5 had the lowest performance for 4 CODs (0.09–0.79), InSilicoVA for 3 CODs (0–0.35), and GPT-3.5 for 1 COD (0.58). All models performed well for injuries, with PCCC values exceeding 0.70 (0.79–0.94), and showed lower performance for malaria (0.35–0.54) and other infections (0.29–0.44). GPT-4 demonstrated an improvement over other models for ill-defined CODs, with improvements greater than 0.30 PCCC (+0.38–0.65), and also showed stronger performance for injuries, with gains of +0.11–0.15 compared to +0.01–0.04 for other models. Performance differences exceeding 0.60 PCCC were observed for epilepsy, leukaemia, other communicable diseases (range: 0.73), ill-defined causes (0.65), and nutritional deficiencies (0.61). In contrast, narrower differences (less than 0.30 PCCC) were seen for malaria (0.20), injuries, and other infections (0.15).

@@

Figure \ref{fig-pccc-cod-neo} shows model performance across 5 neonatal CODs, excluding congenital anomalies (n=2, \textless 1\%) and other causes (n=5, 1\%) due to limited sample sizes. GPT-4 achieved the highest PCCC for 3 of the 5 CODs (0.39–0.71), while GPT-3.5 and InSilicoVA had the highest PCCC for 1 COD each (0.57 and 0.86). GPT-3.5 showed the lowest PCCC for 3 CODs (0–0.13), and InterVA-5 for 2 CODs (0.01 and 0.48). Performance was similar across all models for stillbirths (0.48–0.57 PCCC), though only GPT-4 achieved a PCCC greater than 0 for prematurity-related deaths. InSilicoVA outperformed all other models for neonatal infections, with gains of +0.18–0.73 PCCC. Performance differences greater than 0.6 PCCC were observed for infections (range: 0.73) and prematurity and low birthweight (0.7). Stillbirth showed minimal variation across models (range: 0.09).

@@

This section interprets and contextualizes the findings presented in Section \ref{results}. The comparative advantages and limitations of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA for COD assignment are discussed in Sections \ref{discuss-adv} and \ref{discuss-disadv}, respectively. Study limitations are outlined in Section \ref{discuss-limits}, and directions for future research are presented in Section \ref{discuss-opp}.

@@

This section outlines the strengths of the evaluated models in assigning CODs. Section \ref{discuss-adv-cod} discusses model advantages across specific CODs and age groups. Section \ref{discuss-adv-hscav} highlights the potential for improving efficiency in physician-assisted COD assignment through computational support. Section \ref{discuss-adv-nl} examines the benefits of leveraging natural language narratives in GPT models relative to traditional structured questionnaire data.

@@

At the population level, all models demonstrated comparable performance to physicians (0.74-0.79 CSMF), indicating their potential for estimating COD distributions in large populations. While individual-level performance was lower overall (0.44–0.61 PCCC), several models showed strong performance compared with physicians for specific CODs (up to 0.99 PCCC). GPT-3.5/4 consistently outperformed InSilicoVA and InterVA-5 across most CODs, achieving the highest PCCC for 15 of 17 adult, 7 of 8 child, and 4 of 5 neonatal CODs. In contrast, InSilicoVA showed better performance for select CODs, including road and transport injuries (0.84 PCCC), tuberculosis (0.71), pneumonia (0.78), and neonatal infections (0.86). For CODs where high performance was observed, such as maternal conditions, chronic respiratory diseases, diabetes mellitus, and cancers for GPT-3.5/4 (0.91–0.99 PCCC), and road/transport injuries and neonatal infections for InSilicoVA (0.84 and 0.86 PCCC), the model outputs were more aligned with physician assignment. These findings support the potential utility of combining models based on their strengths for particular CODs. Evaluating performance at the COD level may allow for more targeted deployment of models, maximizing accuracy across disease categories. Table \ref{tab-leadcod} illustrates how different models align with leading CODs identified in prior Sierra Leone studies \citep{carshon-marshChildMaternalAdult2022, ansumanaReportCausesDeath2023}. For example, we may deploy models to estimate asthma and chronic respiratory diseases using GPT-3 (0.94 PCCC), while using GPT-4 and InSilicoVA for diarrhoea and tuberculosis respectively (0.79 and 0.71 PCCC).

@@

Across age groups, all models exhibited a consistent upward trend in performance from 6 months to 14 years, followed by a general decline from ages 15 to 69 years. GPT-3.5/4 outperformed InSilicoVA and InterVA-5 throughout this range, while performance patterns from birth to 5 months were more variable (see Figure \ref{fig-pccc-age}). In adults, performance generally decreased with age, suggesting greater difficulty in assigning CODs among older adults, with a modest improvement observed after age 59. Among children and neonates, performance increased beyond 5 months, indicating greater model reliability as developmental age advanced. Although no model consistently achieved performances greater than 0.8 PCCC in any specific five-year age band, these age-related trends provide valuable insights. Specifically, they align with expectations from clinical literature, where physicians often face greater diagnostic uncertainty in neonatal cases \citep{rasmussenComplexityPhysiciansUnderstanding2019, faisonWhenUnknownUnknowable2023}. The observed patterns underscore the importance of considering developmental stage when interpreting model outputs and comparing them to physician-assigned CODs.
